---
title: "Mapping data in R: `geom_map()` demo"
output: 
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    code_download: true
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(tidyverse)     # for data cleaning and plotting
library(googlesheets4) # for reading googlesheet data
library(lubridate)     # for date manipulation
library(openintro)     # for the abbr2state() function
library(palmerpenguins)# for Palmer penguin data
library(maps)          # for map data
library(ggmap)         # for mapping points on maps
library(ggthemes)      # for more themes (including theme_map())
gs4_deauth()           # To not have to authorize each time you knit.
theme_set(theme_minimal())
```

In this demo, we are going to use `geom_map()` to plot county level COVID 19 cases. Before jumping into the mapping code, we are going to talk about some of the things we did when we read the data in because it is a great review of some of the wrangling tools we've been learning.

We use the `map_data()` function to get MN county border information. We read in a file that gives the population of counties in MN - data are from [here](https://www.minnesota-demographics.com/counties_by_population). And, we read in data from the [NYT](https://github.com/nytimes/covid-19-data) giving current COVID-19 case counts. 

```{r}
county_mn <- map_data("county") %>% 
  # filter to MN
  filter(region == "minnesota") %>% 
  # region is currently the state, so drop that
  select(-region) %>% 
  # rename subregion, region because it will look for that variable
  rename(region = subregion)

# MN county populations
mn_county_pop <- read_csv("https://raw.githubusercontent.com/llendway/mapping/master/data/mn_county_population.csv") %>% 
  # get the county name to match the one from the county_mn file
  # Use separate() to remove "County" at the end of each county name
  separate(county,
           sep = -7:-1,
           into = c("county", "extra")) %>% 
  # Drop that variable 
  select(-extra) %>% 
  # Make counties lower case without any periods
  mutate(county = str_remove(str_to_lower(county), "\\."))


# COVID 19 data from NYT
mn_covid_county <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv") %>% 
  # Filter to MN
  filter(state == "Minnesota") %>% 
  # Keep data from most recent date
  group_by(county, fips) %>% 
  top_n(n = 1, wt = date) %>% 
  # Make counties lower case without any periods
  mutate(county = str_remove(str_to_lower(county), "\\.")) %>% 
  # Join with the county population data
  left_join(mn_county_pop,
            by = "county")
```

Here, I check to make sure I haven't made any mistakes, which I did, at first.

```{r}
mn_covid_county %>% 
  anti_join(distinct_mn_county,
            by = c("county" = "region"))

mn_covid_county %>% 
  anti_join(mn_county_pop,
            by = "county")
```

Now, let's plot the data. First we will plot cases by county.

```{r}
mn_covid_county %>% 
  ggplot() +
  geom_map(map = ,
           aes(map_id = ,
               fill = )) +
  #This assures the map looks decently nice:
  expand_limits(x = , y = ) + 
  theme_map()
```

The previous map mostly reflects the popoulation of the counties. Instead, we can plot the number of cases per 10,000.

```{r}
mn_covid_county %>% 
  ggplot() +
  geom_map(map = county_mn,
           aes(map_id = county,
               fill = )) +
  expand_limits(x = county_mn$long, y = county_mn$lat) +
  theme_map()
```

Lastly, let's make the graph look prettier.

* Add labels  
* Change the fill scale  
* Get rid of legend background  
* Change code chunk options: fig.width and fig.asp

```{r}
mn_covid_county %>% 
  ggplot() +
  geom_map(map = county_mn,
           aes(map_id = county,
               fill = cases/population*10000)) +
  expand_limits(x = county_mn$long, y = county_mn$lat) + 
  labs(title = "",
       caption = "Source: https://github.com/nytimes/covid-19-data",
       fill = "") +
  theme_map()
```
