---
title: "Mapping data in R: `leaflet` demo"
output: 
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    code_download: true
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(tidyverse)     # for data cleaning and plotting
library(googlesheets4) # for reading googlesheet data
library(lubridate)     # for date manipulation
library(openintro)     # for the abbr2state() function
library(palmerpenguins)# for Palmer penguin data
library(maps)          # for map data
library(gplots)        # for col2hex() function
library(ggmap)         # for mapping points on maps
library(RColorBrewer)  # for color palettes
library(sf)            # for working with spatial data
library(leaflet)       # for highly customizable mapping
library(ggthemes)      # for more themes (including theme_map())
gs4_deauth()           # To not have to authorize each time you knit.
theme_set(theme_minimal())
```

```{r data}
# Lisa's Mallorca cycling data
mallorca_bike_day7 <- read_csv("https://www.dropbox.com/s/zc6jan4ltmjtvy0/mallorca_bike_day7.csv?dl=1") %>% 
  select(1:4, speed)

county_mn <- map_data("county") %>% 
  # filter to MN
  filter(region == "minnesota") %>% 
  # region is currently the state, so drop that
  select(-region) %>% 
  # rename subregion, region because it will look for that variable
  rename(region = subregion)

# MN county populations
mn_county_pop <- read_csv("https://raw.githubusercontent.com/llendway/mapping/master/data/mn_county_population.csv") %>% 
  # get the county name to match the one from the county_mn file
  # Use separate() to remove "County" at the end of each county name
  separate(county,
           sep = -7:-1,
           into = c("county", "extra")) %>% 
  # Drop that variable 
  select(-extra) %>% 
  # Make counties lower case without any periods
  mutate(county = str_remove(str_to_lower(county), "\\."))


# COVID 19 data from NYT joined with data above to create MN county-level covid dataset
mn_covid_county <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv") %>% 
  # Filter to MN
  filter(state == "Minnesota") %>% 
  # Keep data from most recent date
  group_by(county, fips) %>% 
  top_n(n = 1, wt = date) %>% 
  # Make counties lower case without any periods
  mutate(county = str_remove(str_to_lower(county), "\\.")) %>% 
  # Join with the county population data
  left_join(mn_county_pop,
            by = "county")
```

This video will show examples of making maps using `leaflet`. I will replicate the plots I created in the demo videos for `ggmap` and `geom_map()`

# Plotting points/markers with `leaflet`

In this part, we will replicate the map of my bike ride in Mallorca. I have already read in the data above. Let's take a look at `mallorca_bike_day7` to remind ourselves what it looks like.

Let's also remember the steps for making a map using `leaflet`:

1. Create a map widget by calling `leaflet()` and telling it the data to use.  
2. Add a base map using `addTiles()` (the default) or `addProviderTiles()`.
3. Add layers to the map by using layer functions (e.g. , `addMarkers()`, `addPolygons()`) to modify the map widget.   
4. Repeat step 3 as desired.  
5. Print the map widget to display it.

## Bare-bones map

Start by just plotting the points (there's a lot of them!). My variable name are not correct although it seems to guess them correclty. Probably better to explicitly name them.

```{r}
leaflet(mallorca_bike_day7) %>% 
  addTiles() %>% 
  addMarkers()
```

## Modify the map

* Change the base map - check out providers preview [here](http://leaflet-extras.github.io/leaflet-providers/preview/.

* Use  `addCircles()` instead of `addMarkers()`.

* Change marker size with `radius`

* Change marker color to color by elevation. This part is tricky because we cannot just map a variable to color. Instead we need to create a variable of colors using the `colorNumeric()` function (show in help).

* Add label of speed (a variable - what do I need?)


```{r}
# create the function to map a variable to color
pal <- colorNumeric("magma",
                    domain = mallorca_bike_day7$ele)

leaflet(mallorca_bike_day7) %>% 
  addProviderTiles(providers$Esri.WorldImagery) %>% 
  addCircles(lng = ~lon,
             lat = ~lat,
             color = ~pal(ele),
             label = ~round(speed,1)) %>% 
  addLegend("topleft",
            pal = pal,
            values = ~ele)
```


# Choropleth map with `leaflet`

In this exercise we will replicate the graph we made that colors MN counties by the number of COVID cases per 10,000 residents.

First, we need to translate the `county_mn` data to an `sf` object. This took me a while to figure out. The final solution was mostly from [here](https://stackoverflow.com/questions/61215968/creating-sf-polygons-from-a-dataframe). We also join on the COVID and population data. 

```{r}
mn_covid_all <- 
  # convert data.frame to sf object
  county_mn %>%
  st_as_sf(coords = c("long", "lat"), crs = "NAD27") %>%
  group_by(region) %>% 
  summarise(geometry = st_combine(geometry)) %>%
  st_cast("POLYGON") %>% 
  # add the covid & population data
  left_join(mn_covid_county,
            by = c("region"="county")) %>% 
  mutate(cases_per_10000 = cases/population*10000)
```

Next, create a map of the counties. Add a label with the county name (use `str_to_title()` so it looks nice.)

```{r}
leaflet(mn_covid_all) %>% 
  addTiles() %>% 
  addPolygons(label = ~str_to_title(region))
```

Let's add some modifications:

* fill by cases_per_10000  
* don't outline the counties (`stroke = FALSE`)  
* change `fillOpacity`  
* change label to have county name and cases_per_10000
* change what happens when you mouse over (`highlightOptions()`)  
* add a legend  

```{r}
pal2 <- colorNumeric("inferno", 
                     domain = mn_covid_all$cases_per_10000)

leaflet(mn_covid_all) %>% 
  addTiles() %>% 
  addPolygons(fillColor = ~pal2(cases_per_10000),
              stroke = FALSE,
              fillOpacity = .7,
              label = ~paste(str_to_title(region),
                                ":",
                                round(cases_per_10000,0),
                                .sep = ""),
              highlight = highlightOptions( 
                                 color = "black",
                                 fillOpacity = 0.9,
                                 bringToFront = FALSE)) %>% 
  addLegend(position = "bottomright",
            pal = pal2, 
            values = ~cases_per_10000,
            title = "Cases per 10,000")
```


