---
title: "Mapping data in R"
output: 
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    code_download: true
---

## Setup

Welcome to another tutorial for this class, COMP/STAT 112: *Introduction to Data Science*! It will be similar to the others, includin demo videos and files embedded in this document and practice problems with hints or solutions at the end.

As most of our files do, we start this one with three R code chunks: 1. options, 2. libraries and settings, 3. data. 

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(tidyverse)     # for data cleaning and plotting
library(googlesheets4) # for reading googlesheet data
library(lubridate)     # for date manipulation
library(palmerpenguins)# for Palmer penguin data
library(maps)          # for map data
library(ggmap)         # for mapping points on maps
gs4_deauth()           # To not have to authorize each time you knit.
```

```{r my_libraries, include=FALSE}
# Lisa needs this, students don't
library(downloadthis) # for including download buttons for files
library(flair) # for highlighting code
```

```{r data}
# Lisa's garden data
garden_harvest <- read_sheet("https://docs.google.com/spreadsheets/d/1DekSazCzKqPS2jnGhKue7tLxRU3GVL1oxi-4bEM5IWw/edit?usp=sharing") %>% 
  mutate(date = ymd(date))

# Seeds/plants (and other garden supply) costs
supply_costs <- read_sheet("https://docs.google.com/spreadsheets/d/1dPVHwZgR9BxpigbHLnA0U99TtVHHQtUzNB9UR0wvb7o/edit?usp=sharing",
  col_types = "ccccnn")

# Planting dates and locations
plant_date_loc <- read_sheet("https://docs.google.com/spreadsheets/d/11YH0NtXQTncQbUse5wOsTtLSKAiNogjUA21jnX5Pnl4/edit?usp=sharing",
  col_types = "cccnDlc")%>% 
  mutate(date = ymd(date))

# Palmer penguins
data("penguins")

# US tuition by state data
us_avg_tuition <- read_csv("https://www.dropbox.com/s/9e0paradcwvuzll/us_avg_tuition.csv?dl=1") %>% 
   mutate(across(starts_with("20"), parse_number))
```

## Learning Goals

After this tutorial, you should be able to do the following:



## Plotting points on a map



