---
title: "Mapping data in R"
output: 
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    code_download: true
---

## Setup

Welcome to another tutorial for this class, COMP/STAT 112: *Introduction to Data Science*! It will be similar to the others, including demo videos and files embedded in this document and practice problems with hints or solutions at the end. There are some new libraries, so be sure to install those first.

As most of our files do, we start this one with three R code chunks: 1. options, 2. libraries and settings, 3. data. 

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(tidyverse)     # for data cleaning and plotting
library(googlesheets4) # for reading googlesheet data
library(lubridate)     # for date manipulation
library(openintro)     # for the abbr2state() function
library(palmerpenguins)# for Palmer penguin data
library(maps)          # for map data
library(ggmap)         # for mapping points on maps
library(ggthemes)      # for more themes (including theme_map())
gs4_deauth()           # To not have to authorize each time you knit.
theme_set(theme_minimal())
```

```{r my_libraries, include=FALSE}
# Lisa needs this, students don't
library(downloadthis) # for including download buttons for files
library(flair) # for highlighting code
```

```{r data}
# Starbucks locations
Starbucks <- read_csv("https://www.macalester.edu/~ajohns24/Data/Starbucks.csv")
```

## Learning Goals

After this tutorial, you should be able to do the following:



## Plotting points on a map

The `Starbucks` data, compiled by Danny Kaplan and provided by Alicia Johnson, contains information about every Starbucks in the world at the time the data were collected. It includes the `Latitude` and `Longitude` of each location.  Let's start by using familiar plotting tools

```{r}
ggplot(data=Starbucks) +
  geom_point(aes(x = Longitude, y = Latitude), 
             alpha = 0.2, 
             size = .1)
```

The point pattern probably looks familiar.  To highlight the geographical nature of this scatterplot, we can superimpose the points on top of a map, using the `ggmap()` function from the `ggmap` library. 

**NOTE**: we used to be able to *easily* bring in Google maps. As of mid-2018, in order to bring those in, you need to have a registered API key. If you want to do that, see `google_key` in the help. Then, see the documentation for `get_map()`. We will bring in other types of maps since Google maps are harder to do now and require you to submit credit card information.

Instead, we bring in a stamen map (there are others you could try, but we'll stick with this). You can also take a look at stamen maps on their [website](http://maps.stamen.com/#watercolor/12/37.7706/-122.3782). First, let's look at an example. 

```{r starbucks-map}
# Get the map information
world <- get_stamenmap(
    bbox = c(left = -180, bottom = -57, right = 179, top = 82.1), 
    maptype = "terrain",
    zoom = 2)

# Plot the points on the map
ggmap(world) + # creates the map "background"
  geom_point(data = Starbucks, 
             aes(x = Longitude, y = Latitude), 
             alpha = .3, 
             size = .1) +
  theme_map()
```

Next, we will walk through the `get_stamenmap()` function arguments. The code below is what was used to get the world map information.

```{r get_stamenmap-code, eval=FALSE}
get_stamenmap(
    bbox = c(left = -180, bottom = -57, right = 179, top = 82.1), 
    maptype = "terrain",
    zoom = 2)
```

**`bbox`**

```{r, echo=FALSE}
decorate_chunk("get_stamenmap-code", eval = FALSE) %>% 
  flair("bbox = ")
```


The `bbox` argument tells it the minimum and maximum latitude and longitude points. So, left is the minimum longitude, right is the maximum longitude, bottom is the minimum latitude, and top is the maximum latitude. I found it helpful to go to [openstreetmap](https://www.openstreetmap.org): zoom in on the area of interest, click export, and you will see all the values you need. I had to modify them slightly, which you can do after your initial plot.

**`maptype`**

```{r, echo=FALSE}
decorate_chunk("get_stamenmap-code", eval = FALSE) %>% 
  flair("maptype = ")
```

The `maptype` tells it the style of the map. Check out the different options by looking in the `get_stamenmap` help (type `?get_stamenmap` in the console).

**`zoom`**

```{r, echo=FALSE}
decorate_chunk("get_stamenmap-code", eval = FALSE) %>% 
  flair("zoom = ")
```

When you make a large area, you need to decrease the zoom, otherwise it will take too long to load. So, it's a good idea to start with a small zoom and you can always make it bigger if you want. This might seem counter-intuitive at first. I think of the zoom level as the level of detail. So, smaller numbers show less detail and larger numbers more detail. I often go to the stamanmaps webpage and search for the location I'm mapping. Then, in the URL, you can see the zoom number. For example, this link  is a map of St. Paul: [http://maps.stamen.com/#terrain/12/44.9531/-93.0904](http://maps.stamen.com/#terrain/12/44.9531/-93.0904). Notice the number `12` next to  `/#terrain/`. That means it is zoomed in at 12. 

**`ggmap()`**

We save the the map information from `get_stamenmap()` to a named value and then use it in `ggmap()`:

```{r, echo=FALSE}
decorate_chunk("starbucks-map", eval = FALSE) %>% 
  flair("world <-") %>% 
  flair("world")
```

The `ggmap()` function will print the "background" map. Think of it as the providing the canvas on which we will plot. This takes the place of our usual `ggplot()`.

```{r}
ggmap(world)
```

After that, we can use the `geom_XXX()` functions from `ggplot2` that we are used to in order to put points, lines, etc. on top of the map. But, we need to remember to also provide the data we are using in the `geom_XXX()` function(s) we use since we do not have the `ggplot()` function in which to provide it. 

```{r, echo=FALSE}
decorate_chunk("starbucks-map", eval = FALSE) %>% 
  flair("data = Starbucks")
```


**`theme_map()`**

The last thing I did in the code was to add `theme_map()`. This is optional, but I often find it makes it look nice.

```{r, echo=FALSE}
decorate_chunk("starbucks-map") %>% 
  flair("theme_map()")
```

So, the final map as a world map as the background with points plotted on top that show the Starbucks locations. The points are .1 of their usual size and have a transparency level of .3. 


### Demo video

### Resources

### Your turn!

#### Exercise 1: More with Starbucks

a. Add an aesthetic to the world map that sets the color of the points according to the ownership type. What, if anything, can you deduce from this visualization?  
b. Construct a new map of Starbucks locations in the Twin Cities metro area (approximately the 5 county metro area).  
c. In the Twin Cities plot, play with the zoom number. What does it do?  (just describe what it does - don't actually include more than one map). 
d. Try a couple different map types (see `get_stamenmap()` in help and look at `maptype`). Include a map with one of the other map types.
e. Add a point to the map that indicates Macalester College and label it appropriately. There are many ways you can do think, but I think it's easiest with the `annotate()` function (see `ggplot2` cheatsheet).

## Choropleths


```{r}
#US states map information - coordinates used to draw borders
states_map <- map_data("state")

#Create a new Starbucks dataset that 
# - filters to the US
# - summarizes the number of Starbucks in each state
# - has full names of states in lowercase letters (to match to states_map)

starbucks_us_by_state <- Starbucks %>% 
  filter(Country == "US") %>% 
  count(`State/Province`) %>% 
  mutate(state_name = str_to_lower(abbr2state(`State/Province`))) 

# map that colors state by number of Starbucks
starbucks_us_by_state %>% 
  ggplot(aes(fill = n)) +
  geom_map(aes(map_id = state_name), map = states_map) +
  #This assures the map looks decently nice:
  expand_limits(x = states_map$long, y = states_map$lat) + 
  theme_map()
```


### Your turn!


## Using `leaflet` to create maps
