---
title: "Mapping data in R"
output: 
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    code_download: true
---

## Setup

Welcome to another tutorial for this class, COMP/STAT 112: *Introduction to Data Science*! It will be similar to the others, including demo videos and files embedded in this document and practice problems with hints or solutions at the end. There are some new libraries, so be sure to install those first.

As most of our files do, we start this one with three R code chunks: 1. options, 2. libraries and settings, 3. data. 

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(tidyverse)     # for data cleaning and plotting
library(googlesheets4) # for reading googlesheet data
library(lubridate)     # for date manipulation
library(openintro)     # for the abbr2state() function
library(palmerpenguins)# for Palmer penguin data
library(maps)          # for map data
library(ggmap)         # for mapping points on maps
library(gplots)        # for col2hex() function
library(RColorBrewer)  # for color palettes
library(sf)            # for working with spatial data
library(leaflet)       # for highly customizable mapping
library(ggthemes)      # for more themes (including theme_map())
gs4_deauth()           # To not have to authorize each time you knit.
theme_set(theme_minimal())
```

```{r my_libraries, include=FALSE}
# Lisa needs this, students don't
library(downloadthis) # for including download buttons for files
library(flair) # for highlighting code
```

```{r data}
# Starbucks locations
Starbucks <- read_csv("https://www.macalester.edu/~ajohns24/Data/Starbucks.csv")
```

## Learning Goals

After this tutorial, you should be able to do the following:

* Plot data points on top of a map using the `ggmap()` function along with `ggplot2` functions.  

* Create choropleth maps using `geom_map()`.  

* Add points and other `ggplot2` features to a map created from `geom_map()`.  

* Understand the basics of creating a map using `leaflet`, including adding points and choropleths to a base map.


## Plotting points on a map

The `Starbucks` data, compiled by Danny Kaplan and provided by Alicia Johnson, contains information about every Starbucks in the world at the time the data were collected. It includes the `Latitude` and `Longitude` of each location.  Let's start by using familiar plotting tools

```{r}
ggplot(data=Starbucks) +
  geom_point(aes(x = Longitude, y = Latitude), 
             alpha = 0.2, 
             size = .1)
```

The point pattern probably looks familiar.  To highlight the geographical nature of this scatterplot, we can superimpose the points on top of a map, using the `ggmap()` function from the `ggmap` library. 

**NOTE**: we used to be able to *easily* bring in Google maps. As of mid-2018, in order to bring those in, you need to have a registered API key. If you want to do that, see `google_key` in the help. Then, see the documentation for `get_map()`. We will bring in other types of maps since Google maps are harder to do now and require you to submit credit card information.

Instead, we bring in a stamen map (there are others you could try, but we'll stick with this). You can also take a look at stamen maps on their [website](http://maps.stamen.com/#watercolor/12/37.7706/-122.3782). First, let's look at an example. 

```{r starbucks-map}
# Get the map information
world <- get_stamenmap(
    bbox = c(left = -180, bottom = -57, right = 179, top = 82.1), 
    maptype = "terrain",
    zoom = 2)

# Plot the points on the map
ggmap(world) + # creates the map "background"
  geom_point(data = Starbucks, 
             aes(x = Longitude, y = Latitude), 
             alpha = .3, 
             size = .1) +
  theme_map()
```

Next, we will walk through the `get_stamenmap()` function arguments. The code below is what was used to get the world map information.

```{r get_stamenmap-code, eval=FALSE}
get_stamenmap(
    bbox = c(left = -180, bottom = -57, right = 179, top = 82.1), 
    maptype = "terrain",
    zoom = 2)
```

**`bbox`**

```{r, echo=FALSE}
decorate_chunk("get_stamenmap-code", eval = FALSE) %>% 
  flair("bbox = ")
```


The `bbox` argument tells it the minimum and maximum latitude and longitude points. So, left is the minimum longitude, right is the maximum longitude, bottom is the minimum latitude, and top is the maximum latitude. I found it helpful to go to [openstreetmap](https://www.openstreetmap.org): zoom in on the area of interest, click export, and you will see all the values you need. I had to modify them slightly, which you can do after your initial plot.

**`maptype`**

```{r, echo=FALSE}
decorate_chunk("get_stamenmap-code", eval = FALSE) %>% 
  flair("maptype = ")
```

The `maptype` tells it the style of the map. Check out the different options by looking in the `get_stamenmap` help (type `?get_stamenmap` in the console).

**`zoom`**

```{r, echo=FALSE}
decorate_chunk("get_stamenmap-code", eval = FALSE) %>% 
  flair("zoom = ")
```

When you make a large area, you need to decrease the zoom, otherwise it will take too long to load. So, it's a good idea to start with a small zoom and you can always make it bigger if you want. This might seem counter-intuitive at first. I think of the zoom level as the level of detail. So, smaller numbers show less detail and larger numbers more detail. I often go to the stamanmaps webpage and search for the location I'm mapping. Then, in the URL, you can see the zoom number. For example, this link  is a map of St. Paul: [http://maps.stamen.com/#terrain/12/44.9531/-93.0904](http://maps.stamen.com/#terrain/12/44.9531/-93.0904). Notice the number `12` next to  `/#terrain/`. That means it is zoomed in at 12. 

**`ggmap()`**

We save the the map information from `get_stamenmap()` to a named value and then use it in `ggmap()`:

```{r, echo=FALSE}
decorate_chunk("starbucks-map", eval = FALSE) %>% 
  flair("world <-") %>% 
  flair("world")
```

The `ggmap()` function will print the "background" map. Think of it as the providing the canvas on which we will plot. This takes the place of our usual `ggplot()`.

```{r}
ggmap(world)
```

After that, we can use the `geom_XXX()` functions from `ggplot2` that we are used to in order to put points, lines, etc. on top of the map. But, we need to remember to also provide the data we are using in the `geom_XXX()` function(s) we use since we do not have the `ggplot()` function in which to provide it. 

```{r, echo=FALSE}
decorate_chunk("starbucks-map", eval = FALSE) %>% 
  flair("data = Starbucks")
```


**`theme_map()`**

The last thing I did in the code was to add `theme_map()`. This is optional, but I often find it makes it look nice.

```{r, echo=FALSE}
decorate_chunk("starbucks-map") %>% 
  flair("theme_map()")
```

So, the final map as a world map as the background with points plotted on top that show the Starbucks locations. The points are .1 of their usual size and have a transparency level of .3. 


### Demo video

With that introduction, you are ready to watch the demo video!

<iframe width="560" height="315" src="https://www.youtube.com/embed/2k8O-Y_uiRU" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

[Voicethread: `ggmap` demo](https://voicethread.com/share/15479688/)

```{r, echo=FALSE}
download_file(
  path = "04_ggmap_demo_no_code.Rmd",
  button_label = "Download ggmap demo file (without code)",
  button_type = "warning",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```

```{r, echo=FALSE}
download_file(
  path = "04_ggmap_demo.Rmd",
  button_label = "Download ggmap demo file (with code)",
  button_type = "info",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```


### Resources

* [Examples](https://github.com/dkahle/ggmap) from `ggmap` maintainer David Kahle  
* [`ggmap` cheatsheet](https://www.nceas.ucsb.edu/sites/default/files/2020-04/ggmapCheatsheet.pdf)

### Your turn!

#### Exercise: More with Starbucks

a. Add an aesthetic to the world map that sets the color of the points according to the ownership type. What, if anything, can you deduce from this visualization?  

b. Construct a new map of Starbucks locations in the Twin Cities metro area (approximately the 5 county metro area).  

c. In the Twin Cities plot, play with the zoom number. What does it do?  (just describe what it does - don't actually include more than one map).  

d. Try a couple different map types (see `get_stamenmap()` in help and look at `maptype`). Include a map with one of the other map types.  

e. Add a point to the map that indicates Macalester College and label it appropriately. There are many ways you can do think, but I think it's easiest with the `annotate()` function (see `ggplot2` cheatsheet).

## Choropleths

Geographical data needn't be expressed by latitude and longitude. For choropleth maps, instead of visualizing our data as points with different aesthetics (size, color, transparency, etc.), we color different regions of the maps based on data values. To do this we need to specify both the geometric regions on which the data resides (counties, states, zip codes, etc.), and then wrangle the data so that there is one value per region. 

Let's return to the Starbucks data. First, we will create a new dataset, `starbucks_us_by_state` that limits the data to the US, finds the number of Starbucks in each state, and creates a state name that is in all lowercase letters that matches the state name in the `region` variable of the `states_map` dataset.

The `states_map` dataset gives information about creating the borders of the US states. The data is retrieved using the `map_data()` function. Run `?map_data` in the console to see more information about what other maps are available. There are also other packages that provide different types of maps.

Then, we can use `geom_map()` to create a choropleth map. Let's take a look at the map and we'll go through the details after.

```{r}
#Create a new Starbucks dataset that 
# - filters to the US
# - summarizes the number of Starbucks in each state
# - has full names of states in lowercase letters (to match to states_map data created next)

starbucks_us_by_state <- Starbucks %>% 
  filter(Country == "US") %>% 
  count(`State/Province`) %>% 
  mutate(state_name = str_to_lower(abbr2state(`State/Province`))) 

#US states map information - coordinates used to draw borders
states_map <- map_data("state")

# map that colors state by number of Starbucks
starbucks_us_by_state %>% 
  ggplot() +
  geom_map(map = states_map,
           aes(map_id = state_name,
               fill = n)) +
  #This assures the map looks decently nice:
  expand_limits(x = states_map$long, y = states_map$lat) + 
  theme_map()
```

Now, let's look more closely at what each piece of the code below is doing.

```{r starbucks-chor-map, eval=FALSE}
starbucks_us_by_state %>% 
  ggplot() +
  geom_map(map = states_map,
           aes(map_id = state_name,
               fill = n)) +
  expand_limits(x = states_map$long, y = states_map$lat) + 
  theme_map()
```

### Choose a map

The `map` argument tells R at which level to create the map. Really, it tells it how to draw all the borders This is a very special data set. According to the `geom_map()` documentation, it is a "data frame that contains the map coordinates ... It **must** contain columns x or long, y or lat, and region or id." We are using the `map_data()` function to create the map file (see above for more detail0. You can open the map data, `states_map`, and see that it adheres to the rules.

```{r, echo=FALSE}
decorate_chunk("starbucks-chor-map", eval = FALSE) %>% 
  flair("map = ")
```

### Connect map id/region variable to data being plotted

The `map_id` inside of `aes()` is a required aesthetic for the `geom_map()` geom. It tells R which variable is the region/id variable, in this case the state. It connects the `region` or `id` from the map (`region` variable in `states_map` dataset, in this example) to the dataset being plotted (`state_name` in `starbucks_us_by_state`, in this example). So `state_name` needs to have the same form as `region`, which is why we modified the state names in `starbucks_us_by_state`.

```{r, echo=FALSE}
decorate_chunk("starbucks-chor-map", eval = FALSE) %>% 
  flair("map_id = ")
```


### Use `ggplot2` features

We tell it to fill in the states by the variable `n`, the number of Starbucks in each state. With the `geom_map()` geom, it will fill in the borders of the regions we defined in the `map` argument.

```{r, echo=FALSE}
decorate_chunk("starbucks-chor-map", eval = FALSE) %>% 
  flair("fill = ")
``` 

### `expand_limits()`

Use `expand_limits()` to assure that the map covers the entire area it's supposed to. We put the longitude variable from `states_map` for the `x` argument and the latitude variable from `states_map` for the `y` argument to assure the map stretches across the entire range of longitudes and latitudes in the map. There may be a better way to do this but I have yet to find it, and when I leave it out, I don't even see the map. 

```{r, echo=FALSE}
decorate_chunk("starbucks-chor-map", eval = FALSE) %>% 
  flair("expand_limits")
```

### `theme_map()`

This is a personal preference. I like the way `theme_map()` makes the map look.

```{r, echo=FALSE}
decorate_chunk("starbucks-chor-map", eval = FALSE) %>% 
  flair("theme_map()")
```

### Add `ggplot2` layers

You can add any of the `ggplot2` layers on top of this map. In this example, I've added the MN Starbucks as points, a title, and changed the legend background (so it doesn't have one and overlap California).

```{r}
starbucks_us_by_state %>% 
  ggplot() +
  geom_map(map = states_map,
           aes(map_id = state_name,
               fill = n)) +
  geom_point(data = Starbucks %>% filter(`State/Province` == "MN"),
             aes(x = Longitude, y = Latitude),
             size = .05,
             alpha = .2, 
             color = "goldenrod") +
  expand_limits(x = states_map$long, y = states_map$lat) + 
  labs(title = "Starbucks in MN") +
  theme_map() +
  theme(legend.background = element_blank())
```

### Demo video

And you are ready to watch another demo video!

<iframe width="560" height="315" src="https://www.youtube.com/embed/iS59a5wDrEM" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

[Voicethread: `geom_map()` demo](https://voicethread.com/share/15484794/)

```{r, echo=FALSE}
download_file(
  path = "04_geom_map_demo_no_code.Rmd",
  button_label = "Download geom_map() demo file (without code)",
  button_type = "warning",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```

```{r, echo=FALSE}
download_file(
  path = "04_geom_map_demo.Rmd",
  button_label = "Download geom_map() demo file (with code)",
  button_type = "info",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```


### Resources

* [ggplot2 documentation](https://ggplot2.tidyverse.org/reference/geom_map.html)  
* [Example](https://rstudio-pubs-static.s3.amazonaws.com/78148_6dd49b5dab4c4f5a8b1a74e5893ff17d.html) by Arie Voorman (some things could be out of date since it's from 2015)

### Your turn!

#### Exercise: Even more with Starbucks

The example I showed did not account for population of each state in the map. In the code below, a new variable is created, `starbucks_per_10000`, that gives the number of Starbucks per 10,000 people. It is in the `starbucks_with_2018_pop_est` dataset.

```{r}
census_pop_est_2018 <- read_csv("https://www.dropbox.com/s/6txwv3b4ng7pepe/us_census_2018_state_pop_est.csv?dl=1") %>% 
  separate(state, into = c("dot","state"), extra = "merge") %>% 
  select(-dot) %>% 
  mutate(state = str_to_lower(state))

starbucks_with_2018_pop_est <-
  starbucks_us_by_state %>% 
  left_join(census_pop_est_2018,
            by = c("state_name" = "state")) %>% 
  mutate(starbucks_per_10000 = (n/est_pop_2018)*10000)
```

a. **`dplyr` review**: Look through the code above and describe what each line of code does.

b. Create a choropleth map that shows the number of Starbucks per 10,000 people on a map of the US. Use a new fill color, add points for all Starbucks in the US (except Hawaii and Alaska), add an informative title for the plot, and include a caption that says who created the plot (you!). Make a conclusion about what you observe.


## Using `leaflet` to create maps
